Determinism, Complexity, and Predictability in Computer Performance

### 2014-01-05 pecomplexity.tex
add stuff about weighted permutation entropy

Utilizing entropy as a measure of temporal complexity is by no means a new idea
\cite{Shannon1951, mantegna1994linguistic}.  Its effective usage
requires categorical data: $x_t \in \mathcal{S}$ for some finite or
countably infinite \emph{alphabet} $\mathcal{S}$, and data taken from
real-world systems is effectively real-valued.  To get around this,
one must discretize the data---typically by binning.  Unfortunately,
this is rarely a good solution to the problem, as the binning of the
values introduces an additional dynamic on top of the intrinsic
dynamics whose entropy is desired.  The field of symbolic dynamics
studies how to discretize a time series in such a way that the
intrinsic behavior is not perverted, but these methods are fragile in
the face of noise and require further understanding of the underlying
system, which defeats the purpose of measuring the entropy in the
first place.

Bandt and Pompe introduced \emph{permutation entropy} (PE)
% also called permutation complexity
as a ``natural complexity measure for time series"
\cite{bandt2002per}.  Permutation entropy employs a method of
discretizing real-valued time series that follows the intrinsic
behavior of the system under examination.  Rather than looking at the
statistics of sequences of values, as is done when computing the
Shannon entropy, permutation entropy looks at the statistics of the
\emph{orderings} of sequences of values using ordinal analysis.
Ordinal analysis of a time series is the process of mapping successive
time-ordered elements of a time series to their value-ordered
permutation of the same size.  By way of example, if $(x_1, x_2, x_3)
= (9, 1, 7)$ then its \emph{ordinal pattern}, $\phi(x_1, x_2, x_3)$,
is $231$ since $x_2 \leq x_3 \leq x_1$.  This method has many
features; among other things, it is robust to noise and requires no
knowledge of the underlying mechanisms.

% It has been shown [amigo?] that the
% normalized permutation entropy \ref{eq:permEntropy} is equal to the
% Kolmogorov-Sinai entropy and therefore is a proxy for the rate at
% which a system becomes unpredictable.

% This spectrum is outlined very nicely in \cite{amigo2012permutation}:
% ``Periodic or quasiperiodic sequences have vanishing or negligible
% complexity. At the opposite end, independent and identically
% distributed random sequences (white noise) have asymptotically
% divergent permutation entropies, owing to the fact that the number of
% allowed (or ``admissible'') ordinal patterns grows superexponentially
% with length. Between both ends lie the kind of sequences we are
% interested in.''

\begin{mydef}[Permutation Entropy]
  Given a time series $\{x_t\}_{t = 1,\dots,T}$. Define
  $\mathcal{S}_n$ as all $n!$ permutations $\pi$ of order $n$. For
  each $\pi \in \mathcal{S}_n$ we determine the relative frequency of
  that permutation occurring in $\{x_t\}_{t = 1,\dots,T}$: $$ p(\pi) =
  \frac{\left|\{t|t\le T-n,\phi(x_{t+1},\dots,x_{t+n}) =
      \pi\}\right|}{T-n+1}$$ Where $|\cdot|$ is set cardinality. The
  \emph{permutation entropy} of order $n\ge 2$ is defined
  as$$H(n) = - \sum_{\pi \in
      \mathcal{S}_n} p(\pi)\log_2p(\pi)$$
\end{mydef}
Notice that $0\le H(n) \le \log_2(n!)$ \cite{bandt2002per}.  With this
in mind, it is common in the literature to normalize permutation
entropy as follows: $\frac{H(n)}{\log_2(n!)}$.  With this convention,
``low'' entropy is close to 0 and ``high'' entropy is close to 1.
Finally, it should be noted that the permutation entropy has been
shown to be identical to the Shannon entropy for many large classes of
systems \cite{amigo2012permutation}.

In practice, calculating permutation entropy involves choosing a good
value for the wordlength $n$. The key consideration
here is that the value be large enough that forbidden ordinals are
discovered, yet small enough that reasonable statistics over the
ordinals are gathered: e.g., $\displaystyle n =
\operatornamewithlimits{argmax}_\ell\{T > 100 \ell!\}$, assuming an
average of 100 counts per ordinal. In the literature, $3\le n \le 6$
is a standard choice---generally without any formal justification. In
theory, the permutation entropy should reach an asymptote with
increasing $n$, but that requires an arbitrarily long time series. In
practice, what one should do is calculate the \emph{persistent}
permutation entropy by increasing $n$ until the result converges, but
data length issues can intrude before that convergence is reached.

% Tables~\ref{tab:TCMpe} \& \ref{tab:IPCpe}

>>>>

Utilizing entropy as a measure of temporal complexity is by no means a new idea
\cite{Shannon1951, mantegna1994linguistic}.  Its effective usage requires
categorical data: $x_t \in \mathcal{S}$ for some finite or countably infinite
\emph{alphabet} $\mathcal{S}$, whereas data taken from real-world systems is
effectively real-valued.  To get around this, one must discretize the data---
typically achieved by binning.  Unfortunately, this is rarely a good solution to
the problem, as the binning of the values introduces an additional dynamic on
top of the intrinsic dynamics whose entropy is desired.  The field of symbolic
dynamics studies how to discretize a time series in such a way that the
intrinsic behavior is not perverted, but these methods are fragile in the face
of noise and require further understanding of the underlying system, which
defeats the purpose of measuring the entropy in the first place.

Bandt and Pompe introduced the \emph{permutation entropy} (PE) as a ``natural
complexity measure for time series" \cite{bandt2002per}.  Permutation entropy
employs a method of discretizing real-valued time series that follows the
intrinsic behavior of the system under examination.  Rather than looking at the
statistics of sequences of values, as is done when computing the Shannon
entropy, permutation entropy looks at the statistics of the \emph{orderings} of
sequences of values using ordinal analysis. Ordinal analysis of a time series is
the process of mapping successive time-ordered elements of a time series to
their value-ordered permutation of the same size.  By way of example, if $(x_1,
x_2, x_3) = (9, 1, 7)$ then its \emph{ordinal pattern}, $\phi(x_1, x_2, x_3)$,
is $231$ since $x_2 \leq x_3 \leq x_1$.  This method has many features; among
other things, it is generally robust to observational noise and requires no
knowledge of the underlying mechanisms.

\begin{mydef}[Permutation Entropy]

  Given a time series $\{x_t\}_{t = 1,\dots,T}$. Define $\mathcal{S}_n$ as all
  $n!$ permutations $\pi$ of order $n$. For each $\pi \in \mathcal{S}_n$ we
  determine the relative frequency of that permutation occurring in $\{x_t\}_{t
  = 1,\dots,T}$:
  \begin{align*}
    p(\pi) = \frac{\left|\{t|t \leq T-n,\phi(x_{t+1},\dots,x_{t+n}) = \pi\}\right|}{T-n+1}
  \end{align*}
  Where $|\cdot|$ is set cardinality. The \emph{permutation entropy} of order $n
  \ge 2$ is defined as
  \begin{align*}
  H(n) = - \sum_{\pi \in \mathcal{S}_n} p(\pi) \log_2 p(\pi)
  \end{align*}

\end{mydef}

Notice that $0\le H(n) \le \log_2(n!)$ \cite{bandt2002per}.  With this in mind,
it is common in the literature to normalize permutation entropy as follows:
$\frac{H(n)}{\log_2(n!)}$.  With this convention, ``low'' entropy is close to 0
and ``high'' entropy is close to 1. Finally, it should be noted that the
permutation entropy has been shown to be identical to the Shannon entropy for
many large classes of systems \cite{amigo2012permutation}.

Here we will be utilizing a variation of the permutation entropy, the
\emph{weighted permutation entropy} (WPE)~\cite{fadlallah2013}. The weighted
permutation entropy attempts to correct for observational noise which is larger
than some trends in the data, but smaller than the larger scale features --- for
example, a signal that switches between two fixed points with noise about those
fixed points. The weighted permutation entropy would be dominated by the
switching rather than by the stochastic fluctuation. To accomplish this, the
\emph{weight} of a permutation is taken into account:
\begin{align*}
  w(x_{t+1:t+n}) = \frac{1}{n}
                 \sum_{x_i \in \{x_{t+1:t+n}\}}
                 \left( x_i - \bar{x}_{t+1:t+n} \right)^2
\end{align*}
where $x_{t+1:t+n}$ is a sequence of values $x_{t+1}, \ldots, x_{t+n}$, and
$\bar{x}_{t+1:t+n}$ is the arithmetic mean of those values.

The weighted probability of a permutation is then:
\begin{align*}
  p_w(\pi) = \frac{\displaystyle \sum_{t \le T - n} w(x_{t+1:t+n}) \cdot \delta(\phi(x_{t:t+n}), \pi) }{\displaystyle \sum_{t \le T - n} w(x_{t+1:t+n})}
\end{align*}
where $\delta(x, y)$ is 1 if $x = y$ and 0 otherwise. Effectively, this weighted
probability enhances permutations involved in ``large'' features and demotes
permutations which are small in amplitude relative to the features of the time
series. The weighted permutation entropy is then:
\begin{align*}
  H_w(n) = - \sum_{\pi \in \mathcal{S}_n} p_w(\pi) \log_2 p_w(\pi)
\end{align*}

In practice, calculating permutation entropy involves choosing a good value for
the wordlength $n$. The key consideration here is that the value be large enough
that forbidden ordinals are discovered, yet small enough that reasonable
statistics over the ordinals are gathered: e.g., $\displaystyle n =
\operatornamewithlimits{argmax}_\ell\{T > 100 \ell!\}$, assuming an average of
100 counts per ordinal. In the literature, $3\le n \le 6$ is a standard choice
---generally without any formal justification. In theory, the permutation
entropy should reach an asymptote with increasing $n$, but that requires an
arbitrarily long time series. In practice, what one should do is calculate the
\emph{persistent} permutation entropy by increasing $n$ until the result
converges, but data length issues can intrude before that convergence is
reached.
